---
title: "Statistics with R"
format: docx
editor: visual
editor_options: 
  chunk_output_type: console
  
knitr:
  opts_chunk:
    #comment: "#>"
    collapse: false
---

# Loading packages 

In this section, we shall load all the necessary packages to enable us download and import the data.

```{r}
#| label: import-package
#| warning: false

if(!require("install.load")){
  install.packages("install.load")
}

install.load::install_load(c("tidyverse", "janitor", "readxl", "openxlsx", "scales"))

theme_set(theme_bw()) # ggplot theme set to theme_bw()
```

# Downloading data into local directory 

We used `download.file()` function to download the data from the [ONS](https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/conceptionandfertilityrates/datasets/conceptionstatisticsenglandandwalesreferencetables/2020/conceptions2020workbook.xlsx) website.  

```{r}
#| label: data-download
#| eval: false

link <- "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/conceptionandfertilityrates/datasets/conceptionstatisticsenglandandwalesreferencetables/2020/conceptions2020workbook.xlsx"

download.file(link, "conceptions2020.xlsx", mode = "wb")
```

# Reading the data

We then import data from the workbook and select the most important variables that would be needed for analysis from the data.

```{r}
#| label: conception-data
#| 
conception <- read.xlsx("conceptions2020.xlsx", sheet = "1a", startRow = 8, sep.names = " ") %>% # Select variables needed for the analysis
  select(
c("Year of conception", "All ages Number of conceptions", "Under 16 Number of conceptions", "Under 18 Number of conceptions", "Under 20 Number of conceptions", "20 to 24 Number of conceptions", "25 to 29 Number of conceptions", "30 to 34 Number of conceptions", "35 to 39 Number of conceptions", "40 and over Number of conceptions"))

```

We can see the information about the data by using:

```{r}
conception %>% 
  dim()

conception %>% 
  names()
```

The trend of conception by year is shown in @fig-trend-line and it can be seen that the rate of conception is gradually decreasing from 2010 till 2020.

```{r}
#| label: fig-trend-line
#| fig-cap: "Number of conceptions by year."
#| fig-width: 6
#| fig-height: 3.5


conception %>% 
  ggplot(aes(x = `Year of conception`, y = `All ages Number of conceptions`))+ geom_line()+scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))+labs( y = "Number of conception", x = "Year")
```


